---
title: 'STAT6984: Homework 2'
author: "Alexander Durbin"
date: "September 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 4
## NOT GETTING .RDATA FILES IN COMMAND LINE BUT SHOULD BE AND WHAT ABOUT DEFAULT ARGUMENTS CONTRASTS

### R Command Line Tools
- R CMD BATCH
    - Called with `R CMD BATCH [options] input_file output_file args`
    - By default produces a `.Rout` file with same name as `input_file`.
    - `.Rout` contains the input and output of running the commands in `input_file` in the `R` console.
    - Plots specified in `input_file` are placed in `Rplots.pdf` on separate pages.
    - Default options are to save `.Rdata` and restore previous `.Rdata` when running scripts. 
    - Does not write output to command line, `cat` arguments in `input_file` appear as console output in `.Rout`.
    - Execution of file is halted when error found in `.R` and no error output appears in console. Warnings are also not printed to console.
    - Can accept conole input using `args` option and `arg <- commandArgs(TRUE)` in `.R` file. Then can index `args` like a vector and assign the values as needed using `var <- eval(parse(text = arg[1]))`. Be clear to not put spaces in `args` as `stdin` uses `space` as a separator.
    
- Rscript
    - Called with `Rscript [options] [-e expr] file args`
    - `expr` is used in absence of `file` and will evaulate the command and print the output to the console.
    - Does not print console output from `file` unless it is a package conflict, warnings, or errors.
    - Also prints plots from `file` on separate pages to `Rplots.pdf`.
    - `args` are similarly used as in `R CMD BATCH` with the appropriate `commandArgs(TRUE)` statement.
    - Does not write to `.Rout` file by default or have options to. If wanting to capture the console output, will need to explicitly write to a file or use log files.
    - Can run single line expressions using `Rscript -e command`.
    - DOES THIS LOAD .RDATA BY DEFAULT?
    
The major difference between these two command line tools is console output. `Rscript` seems more useful for doing quick commands in a remote server whereas `R CMD BATCH` would be more useful for running large programs on a remote server. However, `Rscript` does print warnings and errors and that is incredibly useful when running large jobs, but when combined with log files this could be the superior option.

### Compiling `.Rmd` Files

Assuming the user is in the directory containing the relevant files, the command to compile `.Rmd` files with accompanying `.R` file is

```{r, eval=FALSE}

knitr::knit(input = my_file, tangle = TRUE)

```

The `tangle = TRUE` argument will extract code chunks from the `.Rmd` file and place them in a `.R` file with the same name as `my_file` except if `output_file` was specified for `output` in `knit`.
