---
title: 'Homework 2'
subtitle: 'STAT6984: Advanced Statistical Computing'
author: "Alexander Durbin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# GET HELP FOR UNIX COMMAND ON NUMBER 4, ASK ABOUT PLOT COMMANDS

# Problem 2

The following code defines the original `swap` function. We create a large vector `v` and call `swap` twice. Note that we run the code twice using `system.time()` to account for any background processes that may affect memory allocation, this way we can better identify differences.

```{r}
swap <- function(v, i, j) 
{ 
	tmp <- v[i]
	v[i] <- v[j]
	v[j] <- tmp 
}
v <- 1:1000000000
system.time(swap(v, i=1, j=2))
system.time(swap(v, i=1, j=2))
```

As in the homework prompt, we see that `swap` copies the large vector `v` before swapping the elements. This is an extremely inefficient function.

The following code is replicates switching the first and second elements of `v` using the console.

```{r}

v <- 1:1000000000

system.time({
  tmp <- v[1]
  v[1] <- v[2]
  v[2] <- tmp
})

system.time({
  tmp <- v[1]
  v[1] <- v[2]
  v[2] <- tmp
}) 


```

We see that the second call is instantaneous. This is because `v` is not copied into `swap` before evaluation.

The following code is our implimentation of `swap.eval` using `quote` and `eval`.

```{r}

v <- 1:1000000000
i <- 1
j <- 2

swap.eval <- function(v,...)
{
  eval(
    quote({tmp <- v[i]; v[i] <- v[j]; v[j] <- tmp}), 
    envir=parent.frame()
    )
}

system.time(swap.eval(v, i, j))

system.time(swap.eval(v, i, j))
```

We see that the second call of `swap.eval` is just as efficient as manually using the console, and still a large improvement over `swap`.

The following `swap.do` function uses the `do.call` function, which has options for `quote` and specifying `environment` for arguments. Arguments are entered as a list. `swap.do` is simply `swap` with an added `print` statement.

```{r}

v <- 1:1000000000

swap.do <- function(v, i, j) 
{ 
  tmp <- v[i]
  v[i] <- v[j]
  v[j] <- tmp 
  print(v[1:5])
}

system.time(
  do.call(
    swap.do,
    args = list(v, 1, 2),
    quote = TRUE,
    envir = parent.frame()
  )
)

system.time(
  do.call(
    swap.do,
    args = list(v, 1, 2),
    quote = TRUE,
    envir = parent.frame()
  )
)

```

Surprisingly `swap.do` is just as inefficient as `swap`. This is surprising as the `quote` and `environment` options in `do.call` would lead us to believe that there would be no copying involved. This must not be the case. Additionally, `swap.do` does not actually modify `v` either, this is seen in the consecutive `do.call` statements. Our previous swap functions modified `v` itself, whereas we created a new object here. This may or may not be what we want depending on the situation.

# Problem 3

The following code sources our bisection library for demonstration.

```{r}
  source("bisection.R")
```

## Part A

Bracketing was added to function `bisection` and output was modified in methods `print` and `summary`. Function `bisection` now has a default argument `maxiter=100` that is the maximum iterations allowed when expanding the original interval until $fl \times fr > 0$. New contents added to class `bisection` are `brckt_iter` and `int_prog`. Additionally, `prog` is newly named `root_prog`. `brck_iter` denotes the number of bracket expansion iterations and `int_prog` is similar to `root_prog` in that it is a dataframe of `(xl, xr, fl, fr)` to keep track of the interval expansion progress.

Now, if $fl \times fr > 0$, we iteratively double the interval width before computing $fl, fr$ again and track this progress in `int_prog`. A `stop` command is issued if `maxiter` is exceeded during this process. When $fl \times fr > 0$ is satisfied, the object is updated with `brckt_iter` and `int_prog`. The most recent `xl, xr` are then used as initial entries for the root finding procedure and the rest of the function resembles the original. Appropriate verbose statements were also included to warn the user of interval expansion and output the expansion progress.

The `print` method was simply modified to check if `brckt_iter > 0` before printing appropriate output of the function, root, iterations to find the root, original and expanded interval to find the root, and the number of interval expansion iterations. If there was no interval expansion, the print statement is the same as the original method.

Similarly, the summary procedure is simply modified to check if there was interval expansion before appending the `int_prog` to the output. If there was not any interval expansion, the summary statment is the same as the original method.

Because of the interval expansion procedure, the valid endpoints are used as the initial values for root finding. Additionally, the new valid interval is wider than the user-specified interval. It is because of this that the `plot` method is unchanged.

## Part B

The following code demonstrates the bisection method for the class example with the user-specified interval containing the root.

```{r}

f <- function(x) log(x) - exp(-x)
fr <- bisection(f, 1, 2)
print(fr)
summary(fr)
plot(fr)

```

The following code demonstrates the bisection method with the user-specified interval that does not contain the root.

```{r}

f <- function(x) log(x) - exp(-x)
fr <- bisection(f, 2, 3)
print(fr)
summary(fr)
plot(fr)

```

We see that the appropriate information is printed in the print and summary statements and the initial values for the graph are different from the original.

## Part C

The following code deomonstrates the bisection method for the function $h(x) = (x-1)^3 - 2x^2 + 10 - sin(x)$.

```{r}

f <- function(x) (x-1)^3 - 2*x^2 + 10 - sin(x)
fr <- bisection(f, 1, 2)
print(fr)
summary(fr)
plot(fr)

```

We can further check our function's error handling by testing a function that has no roots. 

```{r, error = TRUE}

f <- function(x) x^2 + 1
fr <- bisection(f, 1, 2)

```

# Problem 4

### R Command Line Tools
- R CMD BATCH
    - Called with `R CMD BATCH [options] input_file output_file args`
    - By default produces a `.Rout` file with same name as `input_file`.
    - `.Rout` contains the input and output of running the commands in `input_file` in the `R` console.
    - Plots specified in `input_file` are placed in `Rplots.pdf` on separate pages.
    - Default options are not to save `.Rdata` and to not restore previous `.Rdata` when running scripts. Will need to manually use the `--save` or `--restore` switches while saving `.Rdata` in your script.
    - Does not write output to command line, `cat` arguments in `input_file` appear as console output in `.Rout`.
    - Execution of file is halted when error found in `.R` and no error output appears in console. Warnings are also not printed to console.
    - Can accept conole input using `args` option and `arg <- commandArgs(TRUE)` in `.R` file. Then can index `args` like a vector and assign the values as needed using `var <- eval(parse(text = arg[1]))`. Be clear to not put spaces in `args` as `stdin` uses `space` as a separator.
    
- Rscript
    - Called with `Rscript [options] [-e expr] file args`
    - `expr` is used in absence of `file` and will evaulate the command and print the output to the console.
    - Does not print console output from `file` unless it is a package conflict, warnings, or error.
    - Also prints plots from `file` on separate pages to `Rplots.pdf`.
    - `args` are similarly used as in `R CMD BATCH` with the appropriate `commandArgs(TRUE)` statement.
    - Does not write to `.Rout` file by default or have options to. If wanting to capture the console output, will need to explicitly write to a file or use log files.
    - Can run single line expressions using `Rscript -e command`.
    - `.Rdata` are not loaded by default.
    
The major difference between these two command line tools is console output. `Rscript` seems more useful for doing quick commands in a remote server whereas `R CMD BATCH` would be more useful for running large programs on a remote server. However, `Rscript` does print warnings and errors and that is incredibly useful when running large jobs, but when combined with log files this could be the superior option to `R CMD BATCH`. Both commands have same default behavior with respect to plotting; Plots are stored in `Rplots.pdf` unless specified otherwise. Additionally, both commands do not load previous `.Rdata` files unless otherwise specified and similarly do not save `.Rdata` upon completion.

### Compiling `.Rmd` Files

Assuming the user is in the directory containing the relevant files, the command to compile `.Rmd` files with accompanying `.R` file is

```{r, eval=FALSE}

knitr::knit(input = my_file, tangle = TRUE)

```

The `tangle = TRUE` argument will extract code chunks from the `.Rmd` file and place them in a `.R` file with the same name as `my_file` except if `output_file` was specified for `output` in the `knit` statement.
