---
title: "Homework 3"
subtitle: "STAT6984: Advanced Statistical Computing"
author: "Alexander Durbin"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r include=FALSE}

library(tidyverse)
library(quantmod)
options("getSymbols.warning4.0" = FALSE)
options("getSymbols.yahoo.warning"=FALSE)

```


# Problem 2

## Part A

`getSymbols` works similarly to `get.quotes` and `get.multiple.quotes` from class but has some slight differences. `getSymbols` *silently* loads the object into the Global Environment. That is, `my_data <- getSymbols(Symbol = 'F')` does not assign `my_data` to the returned value. Instead, `my_data` is assigned the *string*  `'F'` and then an `xts` object is assigned to `F`. This can be worked around by first using `setSymbolLookup(my_data = 'F')` and then calling `getSymbols(Symbol = my_data)`. So we need to use the `auto.assign = FALSE` switch in an assignment statement to get the behavior we expect. Additionally, the `env` option can be used to specify an environment other than the global environment if the user prefers. `getSymbols` is used similarly to our functions from class where `getSymbols(Symbols = 'stock')` or `getSymbols(Symbols = c('stock1', 'stock2', ...,))`. However, providing multiple symbols will load each stock into a separate `xts` object, and not in a single dataframe as in `get.multiple.quotes`. We will then need to `rbind` the various symbols as in `get.multiple.quotes` like in class. Also, `getSymbols` has a `scr = ` argument for specifying the source of the data and defaults to Yahoo!.

__HOW TO HANDLE CRUMB ISSUE?__ The function call uses the command line tool `curl` to connect with Yahoo! and then stores the crumb for the query.

The following code is similar to `get.multiple.quotes` from class with a couple intermediate steps. Each of the returned objects from `getSymbols` is of type `xts`. We convert them to a dataframe first, modify the column names, and add a new "symbol" column.

```{r}

dow30.tickers <-
  c("MMM", "AXP", "AAPL", "BA", "CAT", "CVX","CSCO", "KO", 
    "DIS", "XOM", "GE", "GS", "HD", "IBM", "INTC", "JNJ",
    "JPM", "MCD", "MRK", "MSFT", "NKE", "PFE", "PG", "TRV", 
    "UTX", "UNH", "VZ", "V", "WMT")

getSymbols(Symbols = dow30.tickers)


# for(tck in dow30.tickers){
#   
#   as.name(tck) <- as.data.frame(as.name(tck))
#   colnames(tck) <- stringr::str_extract(colnames(tck), "[^.]*$")
#   tck$symbol <- rep(tck)
#   head(tck)
#   
# }


```