---
title: "Homework 3"
subtitle: "STAT6984: Advanced Statistical Computing"
author: "Alexander Durbin"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r include=FALSE}

library(tidyverse)
library(quantmod)
options("getSymbols.warning4.0" = FALSE)
options("getSymbols.yahoo.warning"=FALSE)
options(stringsAsFactors = FALSE)

```

# ASK ABOUT USING DPLYR AND IF ITS OKAY

# Problem 2

## Part A

`getSymbols` works similarly to `get.quotes` and `get.multiple.quotes` from class but has some slight differences. `getSymbols` *silently* loads the object into the Global Environment. That is, `my_data <- getSymbols(Symbol = 'F')` does not assign `my_data` to the returned value. Instead, `my_data` is assigned the *string*  `'F'` and then an `xts` object is assigned to `F`. This can be worked around by first using `setSymbolLookup(my_data = 'F')` and then calling `getSymbols(Symbol = my_data)`. So we need to use the `auto.assign = FALSE` switch in an assignment statement to get the behavior we expect. Additionally, the `env` option can be used to specify an environment other than the global environment if the user prefers. `getSymbols` is used similarly to our functions from class where `getSymbols(Symbols = 'stock')` or `getSymbols(Symbols = c('stock1', 'stock2', ...,))`. However, providing multiple symbols will load each stock into a separate `xts` object, and not in a single dataframe as in `get.multiple.quotes`. We will then need to `rbind` the various symbols as in `get.multiple.quotes` like in class. Also, `getSymbols` has a `scr = ` argument for specifying the source of the data and defaults to Yahoo!.

The function call uses the command line tool `curl` to connect with Yahoo! and then stores the crumb for the query.

The following function code is similar to `get.multiple.quotes` from class with a couple intermediate steps. Each of the returned objects from `getSymbols` is of type `xts`. We convert them to a dataframe first, modify the column names, and add a new "symbol" column and date column before binding them together.

```{r}

dow30.tickers <-
  c("MMM", "AXP", "AAPL", "BA", "CAT", "CVX","CSCO", "KO", 
    "DIS", "XOM", "GE", "GS", "HD", "IBM", "INTC", "JNJ",
    "JPM", "MCD", "MRK", "MSFT", "NKE", "PFE", "PG", "TRV", 
    "UTX", "UNH", "VZ", "V", "WMT")

my_getSymbols <- function(ticker){
  
  total <- NULL
  
  for (tck in ticker) {
    
    pckg <- getSymbols(Symbols = tck, auto.assign = FALSE)
    tmp <- as.data.frame(pckg, row.names = FALSE)
    tmp$Date <- index(pckg)
    colnames(tmp) <- stringr::str_extract(colnames(tmp), "[^.]*$")
    tmp$Symbol <- tck
    total <- rbind(total, tmp)
    
  }
  
  return(total)
  
}

dow30 <- my_getSymbols(dow30.tickers)

```

## Part B

The following code uses `transform` to compute `Mid`.

```{r}

dow30 <- transform(dow30, Date = as.Date(Date), Mid = (High + Low) / 2)

```

## Part C

The following code defines a new function to calculate returns.

```{r}

calc_returns <- function(your_data, tickers){
  
  new_col <- NULL
  
  for (tck in tickers){
    
    # get data by asset
    tmp <- your_data[your_data$Symbol == tck, ]
    # get the last n-1 values
    new <- tail(tmp, n = -1)
    # get first n-1 values
    old <- head(tmp, n = -1)
    # calculate return
    r <- (new$Mid - old$Mid) / old$Mid
    # make first value NA 
    r <- append(r, NA, after = 0)
    # append to total vector for return
    new_col <- append(new_col, r, after = length(new_col))
    
  }
  
  return(new_col)
  
}

dow30$Return <- calc_returns(dow30, dow30.tickers) 

```

## Part D,E,F ASK

# Problem 3 ASK ABOUT BRK TYPE VARIABLES

```{r, echo = FALSE}

teach <- read.table('teach.csv', sep = ',', header = TRUE)
teach <- teach %>% 
  mutate(marry = ifelse(marry == TRUE, 'married', 'not married')) %>% 
  mutate(degree = as.factor(degree)) %>% 
  mutate(train = ifelse(train == TRUE, 'train', 'not train')) %>% 
  mutate(type = as.factor(type)) %>% 
  mutate(brk = ifelse(brk == TRUE, 'break', 'not break'))

```

The following code provides a plot of `salary` versus `months` by `sex`.

```{r}

ggplot(data = teach) +
  geom_jitter(mapping = aes(x = months, y = salary, color = sex)) +
  ggtitle('Salary Versus Number of Months By Sex')

```

The following code produces several boxplots for our analyses.

```{r}

p <- ggplot(data = teach)
p + geom_boxplot(mapping = aes(sex, salary)) + ggtitle('Salary Distribution by Sex')
p + geom_boxplot(mapping = aes(marry, salary)) + ggtitle('Salary Distribution by Marriage')
p + geom_boxplot(mapping = aes(degree, salary)) + ggtitle('Salary Distribution by Degree')
p + geom_boxplot(mapping = aes(type, salary)) + ggtitle('Salary Distribution by Type')
p + geom_boxplot(mapping = aes(train, salary)) + ggtitle('Salary Distribution by Training')
p + geom_boxplot(mapping = aes(brk, salary)) + ggtitle('Salary Distribution by Break')

```

The following code filters the data by teachers with degree status '1'. We provide a scatterplot and fitted regression lines by sex.

```{r}

teach <- teach %>% 
  filter(degree == 0)

ggplot(data = teach) +
  geom_jitter(mapping = aes(x = months, y = salary, color = sex)) +
  ggtitle('Salary Versus Number of Months By Sex With Degree "0"') +
  geom_smooth(mapping = aes(x = months, y = salary), se = FALSE, method = 'lm')

```

# Problem 4