---
title: "Homework 3"
subtitle: "STAT6984: Advanced Statistical Computing"
author: "Alexander Durbin"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r include=FALSE}

library(tidyverse)
library(quantmod)
options("getSymbols.warning4.0" = FALSE)
options("getSymbols.yahoo.warning"=FALSE)
options(stringsAsFactors = FALSE)

```

# Problem 2

## Part A

`getSymbols` works similarly to `get.quotes` and `get.multiple.quotes` from class but has some slight differences. `getSymbols` *silently* loads the object into the Global Environment. That is, `my_data <- getSymbols(Symbol = 'F')` does not assign `my_data` to the returned value. Instead, `my_data` is assigned the *string*  `'F'` and then an `xts` object is assigned to `F`. This can be worked around by first using `setSymbolLookup(my_data = 'F')` and then calling `getSymbols(Symbol = my_data)`. So we need to use the `auto.assign = FALSE` switch in an assignment statement to get the behavior we expect. Additionally, the `env` option can be used to specify an environment other than the global environment if the user prefers. `getSymbols` is used similarly to our functions from class where `getSymbols(Symbols = 'stock')` or `getSymbols(Symbols = c('stock1', 'stock2', ...,))`. However, providing multiple symbols will load each stock into a separate `xts` object, and not in a single dataframe as in `get.multiple.quotes`. We will then need to `rbind` the various symbols as in `get.multiple.quotes` like in class. Also, `getSymbols` has a `scr = ` argument for specifying the source of the data and defaults to Yahoo!.

The function call uses the command line tool `curl` to connect with Yahoo! and then stores the crumb for the query.

The following function code is similar to `get.multiple.quotes` from class with a couple intermediate steps. Each of the returned objects from `getSymbols` is of type `xts`. We convert them to a dataframe first, modify the column names, and add a new "symbol" column and date column before binding them together.

```{r}

dow30.tickers <-
  c("MMM", "AXP", "AAPL", "BA", "CAT", "CVX","CSCO", "KO", 
    "DIS", "XOM", "GE", "GS", "HD", "IBM", "INTC", "JNJ",
    "JPM", "MCD", "MRK", "MSFT", "NKE", "PFE", "PG", "TRV", 
    "UTX", "UNH", "VZ", "V", "WMT")

my_getSymbols <- function(ticker){
  
  total <- NULL
  
  for (tck in ticker) {
    
    pckg <- getSymbols(Symbols = tck, auto.assign = FALSE)
    tmp <- as.data.frame(pckg, row.names = FALSE)
    tmp$Date <- index(pckg)
    colnames(tmp) <- stringr::str_extract(colnames(tmp), "[^.]*$")
    tmp$Symbol <- tck
    total <- rbind(total, tmp)
    
  }
  
  return(total)
  
}

dow30 <- my_getSymbols(dow30.tickers)

```

## Part B

The following code uses `transform` to compute `Mid`.

```{r}

dow30 <- transform(dow30, Date = as.Date(Date), Mid = (High + Low) / 2)

```

## Part C

The following code defines a new function to calculate returns.

```{r}

calc_returns <- function(your_data, tickers){
  
  new_col <- NULL
  
  for (tck in tickers){
    
    # get data by asset
    tmp <- your_data[your_data$Symbol == tck, ]
    # get the last n-1 values
    new <- tail(tmp, n = -1)
    # get first n-1 values
    old <- head(tmp, n = -1)
    # calculate return
    r <- (new$Mid - old$Mid) / old$Mid
    # make first value NA 
    r <- append(r, NA, after = 0)
    # append to total vector for return
    new_col <- append(new_col, r, after = length(new_col))
    
  }
  
  return(new_col)
  
}

dow30$Return <- calc_returns(dow30, dow30.tickers) 

```

## Part D

The following code calcualtes the Dow Jones Industrial Average (DJIA).

```{r}

ddConst <- 0.14523396877348

dcalc <- function(the_day, the_data){
  
  tmp <- the_data %>% 
    filter(Date == the_day) %>% 
    select(Mid)
  
  pckg <- sum(tmp) / ddConst
  return(pckg)
  
}

dow30 <- dow30 %>% 
  arrange(Date)

djia <- sapply(unique(dow30$Date), function(g) dcalc(g, the_data = dow30))

```

NEED TO COMPARE TO YAHOO DJIA BUT WHICH COLUMN TO COMPARE TO?

## Part E

The following code produces a market-capitalization version of the DJIA.

## Part F

# Problem 3

```{r, echo = FALSE}

teach <- read.table('teach.csv', sep = ',', header = TRUE)
teach <- teach %>% 
  mutate(marry = ifelse(marry == TRUE, 'married', 'not married')) %>% 
  mutate(degree = as.factor(degree)) %>% 
  mutate(train = ifelse(train == TRUE, 'train', 'not train')) %>% 
  mutate(type = as.factor(type)) %>% 
  mutate(brk = ifelse(brk == TRUE, 'break', 'not break'))

```

The following code provides a plot of `salary` versus `months` by `sex`.

```{r}

ggplot(data = teach) + 
  geom_jitter(mapping = aes(x = months, y = salary, color = sex)) +
  ggtitle('Salary Versus Number of Months By Sex')

```

We see that for the first ~100 months, salary increases linearly with months worked. However after the ~100 month mark, there is significant variability in the teacher's salaries. Also it appears that for a given month, male salaries are higher than female salaries. We want to provide appropriate visualizations that show this difference is not just attributed to gender, that the other features of the data are also influencing these differences.

The following code filters out teacher with less than 100 months worked.

```{r}

senior <- teach %>% 
  filter(months > 100)

```

The following code gives a bar chart for males and females based on marriage

```{r}

ggplot(data = teach) +
  geom_bar(mapping = aes(x = sex, fill = marry), position = position_dodge()) +
  ggtitle("Count Of Males and Females By Marriage")

```

We see that there are no married males and about half as many non-married females as married females. There are also more females than males in this subset of the data. It would be interesting to see if marriage is an important factor in female salary. 

The following code gives boxplots for male and female salary filtered by marriage status.

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(sex, salary, fill = marry), position = position_dodge()) +
  ggtitle('Male vs Female Salaries By Marriage Status')

```

We see that non-married females earn more than married females and are similar to the male population. That is, there is no significant difference between non-married females and male salaries and also there appears to be a significant difference in married vs non-married female salary.  

The following code gives bar chart of degree for both sexes. This can be useful to further break down salary average salary differences between the two sexes. 

```{r}

ggplot(data = teach) +
  geom_bar(mapping = aes(x = degree, fill = sex), position = position_dodge()) +
  ggtitle('Count Of Degree By Sex')

```

We see that the majority of the data are of degree '0', with males dominating degrees '1' and '2' with no females of degree '3'.

We next want to compare the salaries of males and females within these groups and the following code does so.

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = sex, y = salary), position = position_dodge()) +
  facet_wrap(~degree, nrow = 2) +
  ggtitle('Distribution of Salary By Sex And Degree')

```

We see that that male salaries of degree '0' are slightly, but no significantly, higher than female degree '0' salaries. The salaries are comparable between sexes for degree type '1' and not comparable for degree type '3'. Most importantly, there appears to be a significant difference between male and female salaries for type '2'.

We also want to consider the `type` variable in our visualizations. The following code provides bar chart for each sex based on `type`. 

```{r}

ggplot(data = teach) +
  geom_bar(mapping = aes(x = type, fill = sex), position = position_dodge()) +
  ggtitle('Counts Of Type By Sex')


```

There appears to be an inverse relationship with `type` and `sex`. There are double female type 'A' than male type 'A' and oppositely for type 'B'. The following code gives boxplots of salary based on sex and type.

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = type, y = salary, fill = sex), position = position_dodge()) +
  ggtitle("Male vs Female Salary By Type")

```

The distributions of type 'A' male and female salaries are comparable and both skewed distributions. However, type 'B' salaries by sex are significantly different, with male salary higher.

We next consider the `train` and `break` variables.

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = train, y = salary, fill = sex), position = position_dodge()) +
  ggtitle("Male vs Female Salary By Train")

```

There is a significant difference in male and female salaries when training. The distributions are comparable with no training, with slight advantage to male teachers. 

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = brk, y = salary, fill = sex), position = position_dodge()) +
  ggtitle("Male vs Female Salary By Break")


```

Salaries of males with breaks are significantly higher than females with breaks. The distributions are comparable with no breaks. 

Thus far, when filtering by single variables `type`, `brk`, and `train`, we have identified significantly higher salaries. We next present boxplots filtered by two variables at a time, similar to sex, salary, and degree type. 

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = sex, y = salary), position = position_dodge()) +
  facet_wrap(~type, nrow = 2) +
  ggtitle('Distribution of Salary By Sex And Type')

```

There appears little difference betwee salaries for type 'A' but a significant difference in salaries for type 'B'.

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = sex, y = salary), position = position_dodge()) +
  facet_wrap(~brk, nrow = 2) +
  ggtitle('Distribution of Salary By Sex And Break')

```

There is a siginificant difference in salaries in males and females who work over summer break. For the population not working over break, males have a slightly higher salary than females, but not significantly. 

```{r}

ggplot(data = teach) +
  geom_boxplot(mapping = aes(x = sex, y = salary), position = position_dodge()) +
  facet_wrap(~train, nrow = 2) +
  ggtitle('Distribution of Salary By Sex And Training')

```

Males and females with no training have comparable salaries. However, there is a significant difference in male and female salaries for those who do train with males having the advantage. 

The above visualizations, while simple, provide insight into what possible factors can influence the salary differences in genders. Making boxplots by both sex and one other factor show that there are possible significant interactions that can account for these differences. 


# Problem 4

For the hockey problem, the following code will execute `hockey.R` and print the standings sorted by conference and division. Teams that have yet to play a single game will not appear in the standings. 

```{r}
source("hockey.R")
```

`nhlstandings.sh` contains the following:

```{r, eval=FALSE}
Rscript -e "source('hockey.R')"
```