---
title: "Homework 1"
author: "Alexander Durbin"
date: "August 30, 2017"
output: 
    pdf_document:
        fig_caption: TRUE
        fig_width: 3
        fig_height: 2
---

# Problem 3

```{r, include = FALSE}
library(tidyverse)
library(tibble)
library(latex2exp)
```


```{r eval = TRUE, echo = FALSE}

count <- 0

my_function <- function(x){
    eval <- 1 - x + 3 * x^2 - x^3
    count <<- count + 1
    return(eval) 
}

raw <- tibble( x = seq(-0.5, 2.5, by = 0.01), y = my_function(x))

ggplot(raw) +
  geom_line(mapping = aes(x, y)) +
  xlab('x') +
  ylab('y') +
  ggtitle(TeX('$f(x) = 1 - x + 3x^2 - x^3$'))



```

Evaluating $f(x) = 1 - x +3x^2 - x^3$ over $[-0.5, 2.5]$, we idenfy critical points
near $0.2$ and $1.8$. Solving for the critical points by hand, we find
$\frac{1}{3}(3 - \sqrt{6}) \approx 0.18$ and $\frac{1}{3}(3+\sqrt{6}) \approx 1.82$
are the critical points of the function. Additionally, we use the `optim` function
to numerically compute the critical points. The results are consistent with our
graphic and the by-hand calculations. `Optim` requires 23 to solve both optimization problems.  

```{r, echo = TRUE, eval = TRUE}

optimise(my_function, interval = c(-0.5, 2.5))
optimise(my_function, interval = c(-0.5, 2.5), maximum = TRUE)
count

```